Finance and Inequality - panel BMA approach\tnoteref{mytitlenote}

### 2020-12-16 Paper/financeinequality.tex
201216 update

\begin{abstract}
We investigate the impact of financial development on income inequality differentiating between depth, efficiency and access to financial markets and institutions. We apply panel Bayesian model averaging framework to address model uncertainty to reveal that financial development has complex relationship   with the income distribution within countries. The access to and efficiency of banking decrease income inequality. The size of the markets has no relevance for overall income inequality, but relates to the higher top income shares. Moreover, unemployment along with investment into non-tangible assets increase income inequality while more redistribution implies lower levels of inequality.
\end{abstract}

@@

\section{Introduction}
\label{ch4sec:intro}
Financial development alters how much the economic opportunities depend on the individual skills, family endowments, social status or political connections as individual may depend on financial system to provide loans to start new business, attain education, or temporarily fund their consumption. The research in the area of financial development and income inequality is well established. \textcite{demirgucc2009finance}, \textcite{claessens2007finance}, and more recently \cite{de2017finance} provide extensive reviews of the topic. A similar theme emerges in all three papers. The implications from theoretical contributions provide conflicting predictions about the relationship and empirical results bring evidence for both positive and negative effect. Although majority of the papers point towards finance tightening the distribution of income this results is not universal with some papers suggesting the opposite while other stress potential non-linearities.

A key divide appears between the effect of financial development on extensive and intensive margin. The extensive margin captures the extend to which individuals, who had not been using financial services before, gain access. On the other hand, the intensive margin describes growing use of finance by the agents who had already been using it before \parencite{demirgucc2009finance}. Financial development on the extensive margin might lead to more equal opportunities and outcomes. Access to credit by previously disadvantaged groups allows human capital accumulation \parencite{galorzeira1993income, galormoav2004, braunetal2019}, formation and growth of new firms \parencite{evans1989estimated,banerjeenewman1990}, with more evenly distributed economic opportunities as a result\footnote{Having similar economic opportunities might decrease the cross-generational inequality, by diminishing the effect of e.g. parental wealth. Depending on the innate abilities and talents of the individuals, however, it may increase the inequality of income within every generation at the same time.}.

On the contrary, intensive margin of financial development might disproportionately benefit the rich who may leverage financial services for their further benefit or to protect their existing rents. \textcite{GreenwoodJovanovic1990} present a model where the finance is the key driver of inequality and the welfare gains accrued by the incumbents - primarily the rich - in the initial development stage. With time, more agents meet the fixed costs of joining the financial intermediaries and they enjoy higher returns. Consequently, the efficiency of resource allocation also increases, which enhances growth and reduces inequality. \textcite{perotti2007investor} present a framework based on political economy. They argument depends on a lobby for lower investor protection to prevent entrance of the new competitors. The politicians require higher bribe from the lobbyist the greater is their accountability for policy decisions. Thus, with increasing accountability, investor protection strengthens and spurs market entry and competition. The authors examine their prediction in a cross-section and show that better investor protection correlates with larger entry rates and higher firm density in more financially intensive sectors\footnote{In addition, they show that the most important factor of accountability is not the formal measure of democratic institutions, but newspaper readership which they interpret as broad awareness of policy choices and their outcomes.}.

Financial development may also have indirect effect on income inequality through economic growth. \textcite{townsendeueda2006} model how finance interacts with production and allocation of credit. If increased use of finance increases the demand for low- relatively to the high-skilled workers, then it may have equalizing consequences for income distribution. Empirical evidence by \textcite{beck2010big} show that bank deregulation and increased competition in loan provision in the \ac{US} primarily benefited the workers with income below the median. Similarly, \textcite{delis2014} provide evidence of bank deregulation and liberalization tightening the income distribution, although this effect is only present in countries with high-quality institutions. They attribute the effect to the changes in labour market conditions and relatively higher wages and working hours of the low-skilled workers following the reforms.

A set of distinct papers explores the relationship between inequality and growth while stressing financial markets imperfections driving the outcomes. Income inequality and growth may intersect through varying channels. Accumulation of savings, unobservable effort, and investment project size favor the prediction of growth inducing inequality. Negative impact of inequality on human capital accumulation, entrepreneurial activity provide argument for the opposing view. 
\textcite{milanovicvan2018inequality} report how income inequality in the \ac{US} has different implications for the future income growth of the rich and the poor. High inequality seems to hurt the prospects of the poor while the top of the distribution is unaffected. The rich thus disproportionately benefit from higher inequality as their subsequent income exhibit faster growth. The authors attribute this effect to the political channel the rich use to lobby in favor of the policies which support their economic interests. Preferences of the rich are ultimately more likely to determine public policy than the preferences of the majority \parencite{gilens_page_2014}. High inequality together with a credit constraint and rich driving the political process results in low government spending and lasting inequality.

The literature does not converge on the conclusions even in the empirical cross-country and panel data studies. The papers link higher levels of financial development with lower levels of inequality \parencite{beck2007finance, hamori2012, gimet2011closer, kunieda2014finance}\footnote{For an extended list, we refer to \textcite{de2017finance}.}. On the other hand, several other estimate a inequality inducing effect of finance \parencite{Jaumotte2013, jauch2016financial, de2017finance}. Finally, some authors claim there the relationship might be non-linear, conditional on a threshold value of financial development \parencite{kim2011nonlinearity,tan2012nonlinear} or institutional quality \parencite{LawSingh2014, delis2014}.

Three papers are the closest to ours, each in a different respect. First, \textcite{de2017finance} examine different dimensions of finance on income inequality. Their results suggest that financial development, financial liberalization, and banking crises all increase pre-tax income inequality within countries. Additionally, they show that the effect of financial liberalization is conditional on democratic accountability. Higher accountability mitigates the impact of liberalization on inequality. On the contrary, the financial development, proxied by the credit to GDP ratio, has inequality increasing effect irrespective of the institutional background. Second, \textcite{naceurzhang2016} take similar approach in considering multiple dimensions --- the access, efficiency, and stability of the financial sector, although not examining the indicators simultaneously.  Third, \textcite{furceri2019robust} apply \ac{WALS} to identify robust determinants of income inequality. Their approach mirrors ours in accounting for model uncertainty in the estimation. Their focus is more general rather than focused primarily on finance. We provide synthesis and extension to these papers in providing more detailed view on the link between finance in shaping income inequality and examining multiple measures of inequality while specifically identifying the determinants of top income shares along with the determinants of the overall income distribution\footnote{Captured by income Gini index.}.

% An exemption from focus solely on the size of financial sector is the study by \textcite{naceurzhang2016} which applies a similar approach to ours, taking into consideration the access, efficiency, and stability of the financial sector. However, there are severe limitation to their study. First, their coverage does not correspond to the availability of the data. Second, they do not account for the different dimensions of finance simultaneously and always use a single indicator of development at a time. Third, they do not explicitly differentiate between the banking sector and financial markets. We provide a more detailed as well as more robust picture of the financial development effect on finance in these dimensions.

% \textcite{de2017finance} examine different dimensions of finance on income inequality. Their results suggest that financial development, financial liberalization, and banking crises all increase market income inequality within countries. Additionally, they show that the effect of financial liberalization is conditional on democratic accountability. The higher the accountability, the less severe is the negative impact of liberalization on inequality. On the contrary, the financial development, proxied by the credit to GDP ratio, has inequality increasing effect irrespective of the institutional background.

% \textcite{furceri2019robust} apply WALS to identify robust determinants of income inequality. They are the closest paper to ours since they account for model uncertainty in the estimation. Their focus if more general rather than on finance specifically. Out work offers more detailed contribution in terms of the role of finance in shaping income inequality. On the top of that, we differ from their analysis by examining multiple measures of inequality and specifically identifying the determinants of top income shares along with the determinants of the overall income distribution represented by income Gini index. We provide further evidence in all these dimensions.

>>>>

\begin{abstract}
  We investigate the impact of financial development on income inequality differentiating between depth, efficiency and access to financial markets and institutions. We apply panel Bayesian model averaging framework to address model uncertainty to reveal that financial development has a complex relationship with the income distribution within countries. The access to and efficiency of banking decrease income inequality. The size of the markets has no relevance for overall income inequality but relates to the higher top income shares. Moreover, unemployment, along with investment into non-tangible assets increase income inequality while more redistribution implies lower levels of inequality.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
Financial development alters how much the economic opportunities depend on the individual skills, family endowments, social status or political connections as individual may depend on financial system to provide loans to start new business, attain education, or temporarily fund their consumption. The research in the area of financial development and income inequality is well established. \textcite{demirgucc2009finance}, \textcite{claessens2007finance}, and more recently \textcite{de2017finance} provide broad reviews of the topic. A similar theme emerges, noting that the implications from theoretical contributions provide conflicting predictions about the relationship. The empirical results then bring evidence for both positive and negative effect. Although the majority of the research points towards finance tightening the distribution of income, this result is not universal with some papers suggesting the opposite while other stress potential non-linearities.

A fundamental divide appears between the effect of financial development on the extensive and intensive margin. The extensive margin captures the extent to which individuals, who had not been using financial services before, gain access. On the other hand, the intensive margin describes the growing use of finance by the agents who had already been using it before \parencite{demirgucc2009finance}. Financial development on the extensive margin might lead to more equal opportunities and outcomes. Access to credit by previously disadvantaged groups allows human capital accumulation \parencite{galorzeira1993income, galormoav2004, braunetal2019} as well as formation and growth of new firms \parencite{evans1989estimated, banerjeenewman1990}, with more evenly distributed economic opportunities as a result\footnote{Having similar economic opportunities might decrease the cross-generational inequality, by diminishing the effect of, e.g. parental wealth. Depending on the innate abilities and talents of the individuals, however, it may increase the inequality of income within every generation at the same time.}.

On the contrary, the intensive margin of financial development might disproportionately benefit the rich who may leverage financial services for their further benefit or to protect their existing rents. \textcite{GreenwoodJovanovic1990} present a model where the finance is the key driver of inequality and the welfare gains accrued by the incumbents - primarily the rich - in the initial development stage. With time, more agents meet the fixed costs of joining the financial intermediaries, and they enjoy higher returns. Consequently, the efficiency of resource allocation also increases, which enhances growth and reduces inequality. \textcite{perotti2007investor} present a framework based on political economy. Their argument depends on a lobby for lower investor protection to prevent entrance of the new competitors. The politicians require higher bribe from the lobbyist the greater is their accountability for policy decisions. Thus, with increasing accountability, investor protection strengthens and spurs market entry and competition. The authors examine their prediction in a cross-section and show that better investor protection correlates with larger entry rates and higher firm density in more financially intensive sectors\footnote{In addition, they demonstrate that the most important factor of accountability is not the formal measure of democratic institutions, but newspaper readership which they interpret as broad awareness of policy choices and their outcomes.}.

Financial development may also have an indirect effect on income inequality through economic growth. \textcite{townsendeueda2006} model how finance interacts with production and allocation of credit. If increased use of finance increases the demand for low- relatively to the high-skilled workers, then it may have equalizing consequences for income distribution. Empirical evidence by \textcite{beck2010big} shows that bank deregulation and increased competition in loan provision in the US primarily benefited the workers with income below the median. Similarly, \textcite{delis2014} provide evidence of bank deregulation and liberalization tightening the income distribution, although this effect is only present in countries with high-quality institutions. They attribute the effect to the changes in labour market conditions and relatively higher wages and working hours of the low-skilled workers following the reforms.

A set of distinct papers explores the relationship between inequality and growth while stressing financial market imperfections in driving the outcomes. Income inequality and growth may intersect through varying channels. Accumulation of savings, unobservable effort, and investment project size favour the prediction of growth-inducing inequality. The negative impact of inequality on human capital accumulation and entrepreneurial activity provides an argument for the opposing view. 
\textcite{milanovicvan2018inequality} report how income inequality in the US has different implications for the future income growth of the rich and the poor. High inequality seems to hurt the prospects of the poor while the top of the distribution is unaffected. The rich thus disproportionately benefit from higher inequality as their subsequent income exhibits faster growth. The authors attribute this effect to the political channel where the rich use to lobby in favor of the policies which support their economic interests. Preferences of the rich are ultimately more likely to determine public policy than the preferences of the majority \parencite{gilens_page_2014}. High inequality together with a credit constraint and rich driving the political process results in low government spending and lasting inequality.

The literature does not converge on the conclusions even in the empirical cross-country and panel data studies. The papers link higher levels of financial development with lower levels of inequality \parencite{beck2007finance, hamori2012, gimet2011closer, kunieda2014finance}\footnote{For an extended list, we refer to \textcite{de2017finance}.}. On the other hand, several other estimate an inequality inducing effect of finance \parencite{Jaumotte2013, jauch2016financial, de2017finance}. Finally, some authors claim that the relationship might be non-linear, conditional on a threshold value of financial development \parencite{kim2011nonlinearity,tan2012nonlinear} or institutional quality \parencite{LawSingh2014, delis2014}.

The contribution of the paper is fourfold. First, we efficiently account for model uncertainty relying on the panel \ac{BMA} framework. Second, we use the \ac{WID} data on the top income shares, collected based on the tax collection data. Arguably, the data is superior to the income surveys as it amends issues of underrepresentation of high-income individuals and underreporting income. Third, we simultaneously consider different proxies of financial development to identify the most relevant channels through which finance affects inequality and, fourth, we examine multiple measures of income inequality to distinguish the diverse effects of finance across the income distribution.

Three papers are the closest to ours, each in a different respects. First, \textcite{de2017finance} examine different dimensions of finance on income inequality. Their results suggest that financial development, financial liberalization, and banking crises all increase pre-tax income inequality within countries. Additionally, they show that the effect of financial liberalization is conditional on democratic accountability. Higher accountability mitigates the impact of liberalization on inequality. On the contrary, the financial development, proxied by the credit to GDP ratio, has inequality increasing effect irrespective of the institutional background. Second, \textcite{naceurzhang2016} take a similar approach in considering multiple dimensions --- the access, efficiency, and stability --- of the financial sector when studying determinants of inequality and poverty, but they do not attempt to consider the indicators simultaneously.  Third, \textcite{furceri2019robust} apply \ac{WALS} to identify robust determinants of income inequality. Their approach mirrors ours in accounting for model uncertainty in the estimation, but their focus is more general rather than focused primarily on finance. We provide a synthesis and a natural extension to these papers by offering more detail on how finance in shapes income inequality. On top of that, we examine multiple measures of inequality while specifically identifying the determinants of top income shares along with the determinants of the overall income distribution.

The chapter continues with a description of the data and methodology in Section \ref{sec:datameth}. We then provide two sets of results in Section \ref{sec:results}. We employ the Gini coefficient as a measure of inequality in one and income shares of top decile and percentile in the other. Section \ref{sec:robustness} provides robustness checks of the results and Section \ref{sec:conclusion} concludes.

%
%

\section{Data and methodology}
\label{sec:datameth}
The key variable in the paper is the measure of income inequality. We want to examine how financial development affects income inequality and whether the effect might be different at the top quantiles of the income distribution. As the overall measure of income inequality, we rely upon the after-tax Gini coefficient from \ac{SWIID} by \textcite{Solt2019}, which is a standard resource in the literature\footnote{There are alternative sources of for Gini coefficient, e.g. \ac{WIID} or \ac{LIS}, but each of them brings limitations in terms of comparability or coverage.}. Its critical advantage lies in the widespread coverage across countries and time and a unified methodology which provides a reasonable level of comparability. It typically takes values in the interval between 0 and 100 where the former suggests perfect equality (everyone in the economy enjoys the same income) and the latter perfect inequality (all the income goes to only a single unit). We depart from existing papers slightly in considering the after-tax rather than the before-tax income distribution as the dependent variable. We choose the after-tax income Gini coefficient as we also include the proxy of redistribution among the regressors to account for taxation and transfers indirectly. Since we define redistribution as a difference between before-tax and after-tax Gini coefficients, the estimate is not substantially influenced by using either of the two as the dependent variable\footnote{\textcite{furceri2019robust} make a similar argument that unless redistribution is systematically correlated with other regressors, their effect on the net and gross inequality should be same. As a robustness check, we also ran the estimation with before-tax Gini coefficient, and the single qualitative difference is only the sign of the posterior mean of redistribution coefficient.}. In our case, using after-tax allows for convenient interpretation.

To explore the relationship in the top part of the distribution, we choose top income shares from the \ac{WID}\footnote{The methodology and guidelines to database are provided by \textcite{alvaredo2016distributional}.}. The surveys suffer from well-known issues of underrepresentation of the top income earners and the distortions resulting from the self-reported character of the data. This can influence not only the top income shares originating from the survey data, but it may also distort the overall measures of inequality. The data in \ac{WID} make use of income tax records in individual countries and the derived shares obtained using the consistent methodology of \ac{DINA} are arguably more reliable relative to the survey-based measures which are the primary source of majority estimates of income distributions.

The data spans from 2000 to 2014. We follow the literature \parencite{dabla2015causes,de2017finance} and average both the inequality measure (dependent variable) and the potential determinants (independent variables) across 3-year intervals. There are important reasons for looking at the averages than observation in individual years. Annual macroeconomic data are subject to fluctuations, and the data on income inequality is noisy \parencite{delis2014}. Averaging should diminish the level of noise. On the top of that, the variables at the center of our analysis, e.g. stock market capitalization or credit to \ac{GDP}, are likely to be affected by the business cycles and volatile on a yearly basis. A similar argument holds for top income shares, as they depend, among other things, on the bonuses paid out each year as well as capital income. We want to explore the long-term rather than the short-term relationship and that guides the choice of averaged data. Faced against the trade-off between the length of the averaging periods and available observations in the time dimension, we take a compromise of three years in contrast to the literature, where the 5-year intervals generally apply. The availability of financial development indicators limits the analysis to a period from 2000 onward, and we prefer to keep at least five unique time periods to just three under the case of 5-year average\footnote{Nevertheless, we run the estimation with 5-year averages of data as a robustness check and find no critical qualitative differences compared to the baseline. The results are available in the Appendix of the paper}. Two reasons further support this decision. First, the 3-year models display much better convergence which we grant to more available observations and higher variation in the 3-year averaged data. Second, more periods allow for a more robust estimate when we lag the explanatory variables in order to address endogeneity in one of the robustness checks.

Table \ref{tab:sum} reports the summary statistics of the income inequality variables and financial development indicators.

\begin{table}[ht!]
  \small
  \caption{Summary statistics of selected variables}
  \label{tab:sum}
  \centering
     \begin{tabular}{lrrrr}
      \toprule
      Variable & Mean & St. Dev. & Min & Max \\
      \midrule
      After-tax Gini index  & 36.38 & 8.12 & 22.88 & 61.16 \\
      Top 10\% income share & 0.42 & 0.12 & 0.24 & 0.71 \\
      Top 1\% income share & 0.14 & 0.06 & 0.05 & 0.38 \\
      FIA & 0.42 & 0.32 & 0.01 & 1.00 \\
      FIE & 0.60 & 0.13 & 0.11 & 0.81 \\
      FID & 0.37 & 0.30 & 0.01 & 1.00 \\      
      FMD & 0.33 & 0.32 & 0.00 & 0.99 \\
      \bottomrule
  \end{tabular}
\end{table}
%
%
%
%
%

\begin{table}[ht!]
  \small
  \caption{Correlation matrix of selected variables}
  \label{tab:corr}
  \centering
  \begin{tabular}{lrrrrrrr}
    \toprule
  %  & GiniNet & Top10share & Top1share & FIA & FIE & FID & FMD \\
    % \midrule
    After-tax Gini index  & . &  &  &  &  &  &  \\
    Top 10\% income share & 0.47 & . &  &  &  &  &  \\ 
    Top 1\% income share & 0.39 & 0.84 & . &  &  &  &  \\
    FIA & -0.14 & -0.09 & -0.12 & . &  &  &  \\
    FIE & -0.14 & -0.10 & -0.06 & 0.28 & . &  &  \\ 
    FID & -0.07 & 0.12 & 0.06 & 0.60 & 0.24 & . &  \\
    FMD & 0.03 & 0.2 & 0.13 & 0.36 & 0.1 & 0.40 & . \\ 
     \bottomrule
  \end{tabular}
  \end{table}

  We obtain the financial development indicators from \ac{GFDD}. The database offers detailed indicators along four dimensions of financial systems and allows to estimate the effect of changes in access, size, efficiency, and stability of financial markets. Furthermore, we can distinguish between the banking sector and financial markets in all these dimensions. The data for access and stability of stock markets remains sparse in the concerned period we must leave them out of the analysis. We use the version of financial indicators from \textcite{svirydzenka2016introducing}. The authors make use of principal component analysis in order to construct aggregate indicators in each characteristic of the financial sector. In summary, we have indicators of financial institutions depth (FID), financial markets depth (FMD), access to financial institutions (FIA), the efficiency of financial markets (FME), and institutions (FIE)\footnote{\textcite{svirydzenka2016introducing} extrapolate the indicators from top to bottom if the original variables are unavailable, we make sure that that at least one variable is available for the construction of the index and no artificial correlation introduced to the data.}. We report the composition of each indicator in Table \ref{tab:finind}.

\begin{figure}[ht!]
  \caption{Gini Coefficient and top shares}
  \label{fig:gini_topshares}
  \centering
  \includegraphics[width=0.8\textwidth, keepaspectratio]{figures//plots_ineq}
\end{figure}

% Figure \ref{fig:gini_findev} shows the scatter plots of all the available observations after averaging between Gini coefficient and selected financial indicators. Although the data is noisy, it suggest negative correlation between the variables. Figure \ref{fig:gini_findev_dm} then shows the corresponding scatter plots of the data which was demeaned using the within country averages over time. %The relationship between Gini and financial indicators appears considerably weaker following the transformation, with the exception of access to financial institutions where visual inspection allows pointing to a negative correlation.

We base the choice of other explanatory variables on the reviews of income inequality drivers \parencite{roineetal2009,nolan2019drivers}, a related study of finance-inequality nexus \parencite{de2017finance}, and a more general inquiry into the robust determinants of income inequality \parencite{furceri2019robust}. The potential regressors could be categorized into several groups. They control for economic and financial development, demographics, globalization, and institutional background. Table \ref{app:exog} reports all the control variables and their sources. 
%
%
% \section{Methodology}, could be extended if necessary
% \label{sec:methodology}
Methodologically, we rely on the \ac{BMA} approach, which conveniently addresses the issue of model uncertainty with a large number of potential regressors. The advantages and statistical properties of the \ac{BMA} have been described in \textcite{Koop2003}. In application to the panel data, we make use of Frisch-Waugh-Lovell theorem and demean all the variables using time averages for individual observations. Using the time-demeaned observations in the estimation is equivalent to the estimate using the dummy variables for individual cross-sectional units. The key assumptions in \ac{BMA} are on parameter and model priors. For the parameter prior, we turn to so-called hyper-g prior. The prior provides more robust results than some other traditionally applied $g$ priors such as \ac{UIP}, \ac{BRIC} \parencite{feldkircher2012impact}. As for the model priors, the baseline estimate relies on the uniform model prior. We choose the model before to remain agnostic about the prior probability of each examined model. While uniform prior assigns the same prior probability to each model, the distribution of the prior model space is concentrated around $k/2$, where $k$ is the number of potential covariates, and it may consequentially gravitate towards larger model sizes and a higher number of covariates\footnote{See \textcite{LeySteel2009} for details.}.

\section{Results}
\label{sec:results}
We examine the determinants of inequality in the panel \ac{BMA} framework and present the results in the following sections. We start with a model where we capture the overall inequality by Gini index in Subsection \ref{subsec:gini}. We then continue to estimations where we consider the shares of income going to the top 10\% and top 1\% of the income distribution as our dependent variable. We check the robustness of our estimates by employing alternative model and parameter priors throughout the analysis.

\subsection{Gini index of income distribution}\label{subsec:gini}
We focus the analysis on the relationship between the indicators representing various aspects of financial development and income inequality. Figure \ref{fig:gini_findev_dm} outlines the expected link after we have demeaned the variables using the cross-sectional averages. The relationship is not particularly strong, but we observe a negative correlation between Gini index and indicators of access and efficiency of financial institutions, as suggested by a linear estimate. For size indicators of financial market and institutions, the link appears much weaker.

\begin{figure}[ht!]
  \caption{Gini Coefficient and Financial Development Indicators}
  \label{fig:gini_findev_dm}
  \centering
  \includegraphics[width=0.8\textwidth, keepaspectratio]{figures//plots_findev_gini_dm}
\end{figure}

% \begin{figure}
%   \caption{Gini Coefficient and Financial Development Indicators}
%   \label{fig:gini_findev}
%   \includegraphics[width=\textwidth, keepaspectratio]{figures//plots_findev_gini}
% \end{figure}

Table \ref{res:baseline_gini} reports the baseline results. Overall, we have 16 variables with \ac{PIP} above 0.8. The number of unique relevant regressors effectively shrinks by two if we abstract from the quadratic terms of the \ac{GDP} per capita and the education index. Most of the estimated posterior means exhibit expected signs.

The only financial indicators which occur among the top regressors are access and efficiency of financial institutions with \acp{PIP} of 1 and 0.88, respectively. The posterior mean on the coefficients in both dimensions is negative, so higher levels of access and efficiency are associated with lower levels of income inequality. The inequality decreasing effect of access to finance on inequality mirrors \textcite{hasan2020finance} for wealth inequality, and partially also \textcite{furceri2019robust, naceurzhang2016} who document similar effect. The observation on inequality decreasing effect of access to finance also supports \textcite{claessens2007finance} who suggest that access may equalize economic opportunities and lead to a more evenly distributed income as well as with theoretical predictions \parencite{braunetal2019,galormoav2004,banerjeenewman1990}. The efficiency of financial intermediation putting downward pressure on income inequality also has a precedent in \textcite{gimet2011closer}\footnote{The authors measure the efficiency of the banking sector by the difference between the lending rate and the deposit rate (spread). They argue that higher spread reflects low competition and high transaction costs. The imperfections in the credit market can skew the credit to high-income, rich households who can provide significant collateral, reinforcing the existing inequality.}. We fail to confirm that the efficiency of financial institutions is a robust determinant of inequality; however, as the \ac{PIP} markedly decreases under alternative model priors. None of the size indicators of financial institutions or markets has a high probability of inclusion.

Education expenditures (\% share of \ac{GDP}) along with the education index calculated using mean and expected years of schooling show inequality decreasing effect. It is in line with the prediction of \textcite{goldin2009race,deaton2013great} who claim that the skill-biased technological change should be mitigated by education. \textcite{oecd2011divided} finds similar evidence for a panel of advanced countries while \textcite{furceri2019robust} also suggest a negative relationship, although it is not entirely robust to variable and sample selection. The effect of education diminishes at the higher levels of schooling as the quadratic term has also high \ac{PIP} and positive posterior mean.

\begin{table}[ht!]
  \caption{BMA, baseline results. Dependent variable after-tax Gini index, 394 observations.}\label{res:baseline_gini}
  \footnotesize
  \centering
  \begin{tabular}{lrrr}
    \toprule
                           & PIP & Post Mean & Post SD \\
    \midrule
  Education expenditures & 1.00 & -0.14506 & 0.05042 \\
  GDP per capita & 1.00 & 1.62811 & 0.56870 \\
  Unemployment & 1.00 & 0.23629 & 0.05907 \\
  Non-equipment investment & 1.00 & 0.14977 & 0.05285 \\ 
  Access to financial institutions & 1.00 & -0.24629 & 0.06342 \\
  Education index (UN) & 1.00 & -0.58853 & 0.23865 \\ 
  Redistribution & 1.00 & -0.22055 & 0.05051 \\
  Education index sq. & 1.00 & 0.82252 & 0.22008 \\
  GDP per capita sq. & 1.00 & -1.44906 & 0.57565 \\
  Life expectancy & 0.99 & -0.22879 & 0.09750 \\ 
  Economic freedom & 0.99 & 0.18518 & 0.06941 \\
  Value added in agriculture & 0.97 & -0.12361 & 0.05947 \\
  Government expenditures & 0.95 & 0.12259 & 0.05743 \\ 
  Total population & 0.95 & -0.15378 & 0.08478 \\
  Financial institutions efficiency & 0.88 & -0.08374 & 0.05599 \\
  Inflation & 0.86 & 0.16879 & 0.12487 \\
  \midrule
  Inflation sq. & 0.66 & -0.11127 & 0.11201 \\ 
  Value added in industry & 0.62 & -0.05825 & 0.06407 \\
  Equipment investment & 0.44 & -0.03383 & 0.05359 \\ 
  Financial markets depth & 0.27 & 0.01554 & 0.03673 \\ 
  Gross domestic savings & 0.27 & -0.02022 & 0.04597 \\
  % Period 4 & 0.25 & 0.01282 & 0.03330 \\ 
  Social globalization & 0.23 & 0.02365 & 0.06279 \\
  Restrictions on globalization & 0.15 & 0.00874 & 0.03214 \\
  Trade openness & 0.13 & 0.00615 & 0.02589 \\
  Left-wing orientation & 0.11 & 0.00339 & 0.01720 \\ 
  Rule of law & 0.10 & -0.00333 & 0.01812 \\ 
  Net FDI (\% GDP) & 0.10 & -0.00283 & 0.01575 \\ 
  Natural resources rents & 0.08 & -0.00223 & 0.01583 \\ 
  Chinn-Ito index & 0.07 & -0.00205 & 0.01780 \\ 
  % Period 5 & 0.07 & 0.00025 & 0.02188 \\ 
  % Period 2 & 0.07 & 0.00133 & 0.01321 \\
  % Period 3 & 0.06 & -0.00027 & 0.01423 \\
  Financial globalization & 0.06 & -0.00134 & 0.01466 \\
  Civil liberties \& political rights & 0.06 & 0.00075 & 0.01131 \\ 
  Financial institutions depth & 0.05 & 0.00060 & 0.01316 \\ 
  GDP growth & 0.05 & 0.00016 & 0.01255 \\
  Population growth & 0.05 & 0.00055 & 0.01052 \\ 
  Political globalization & 0.05 & 0.00000 & 0.01432 \\  
    \bottomrule
  \end{tabular}
  \end{table}

  We find evidence for the Kuznets' hypothesis about inequality and economic development. We include three regressors which primarily capture the level of development - GDP per capita along with its square term, the share of value-added in agriculture, and share of value-added in industry. Exploring the results of the baseline, We find evidence for an inverted-U curve for GDP per capita and baseline \acp{PIP} for the linear and square component. This suggests that inequality tends to be lower at the initial stages of economic development, then increases as the economic output grows, and only starts decreasing after reaching a threshold. The \ac{PIP} of 0.97 for value-added in agriculture along with a negative posterior mean for its coefficient supports this idea further. The economies at lower stages of development generally report higher shares of the agricultural sector which is also more labour intensive, i.e. not exacerbating the inequalities that stem from an unequal distribution of capital. The level of unemployment is associated with elevated income inequality. The mechanism is direct through the lower share of income going to labour in case of high unemployment rates. The effect of unemployment is also documented on cross-section by \textcite{furceri2019robust}. We also explore the effect of non-equipment investment\footnote{We construct the indicator using the detail on capital investment prom \ac{PWT} and split the overall investment into non-equipment (structures, transportation, and other assets - software / intellectual property products) and equipment investment (machinery, computers + communication equipment).}, which we believe could proxy for the technological progress and skill-biased technological change. The \ac{PIP} suggests positive correlation with income inequality as \textcite{goldin2009race,dabla2015causes}.

  We rely on redistribution to capture the effect of taxes and social expenditure on income inequality. Our measure is the difference between before-tax and after-tax Gini coefficient from the \ac{SWIID} database. Therefore, we concentrate on the direct effects of policies on household income. Given the global nature of our data and its limitations, we cannot estimate the potential indirect effects on the pre-tax distribution of income using, for example, top marginal tax rates \parencite{alvaredoetal2013} nor corporate tax rates \parencite{fuest2018higher}. Nevertheless, the \ac{PIP} of our indicator is very high and it negatively relates to income inequality as expected. The government expenditures, an often used regressor in the literature, also has a perfect inclusion probability, but with positive posterior mean. The intuitive first-order effect of government expenditure should be through reduction of inequality through general social spending on transfers, education, and health. We account for these explicitly, however, with redistribution and education expenditures of the government. \textcite{anderson2017does} introduce a meta-analysis where they show how the examined relationship depends on the type of spending considered and the measure of income inequality. Additionally, they suggest that the redistributive impact has not extended over the entire distribution, but has mostly centred towards middle-income groups. In line with their conclusions, our results suggest the effect of government spending might run in a positive direction when the key equalizing function have been accounted for or perhaps point towards reverse causality.

  Life expectancy should proxy for changing demographics. We find a negative link between life expectancy and income inequality. \textcite{goldsteinlee2014} examine the channels between population ageing and inequality and argue that stretching the economic life-cycle is associated with larger within-group variance as cohort ages. They claim this should pronounce inequality in older populations. On the other hand, if we consider retirement age, pension structure in many countries equalize income flows and therefore the older populations may report lower levels of inequality. The posterior mean of our estimated coefficient indicates the latter scenario.

  Inflation is the only variable which could proxy for monetary policy and its influence on income inequality. It is among the relevant regressors with \ac{PIP} of 0.86 and shows inequality inducing association. The interest in the effects of monetary policy, and macroprudential tools in particular, on inequality is relatively recent, following the introduction of unconventional measures following the Great Recession \parencite{frost2018macroprudential}. Theoretically, the effect is ambiguous as high inflation of asset prices might benefit high-income households more as they receive higher shares of financial income. Also, the assets held by low-income households tend to be much more liquid; thus inflation hurts them more relatively to the high-income ones. On the contrary, cut in rates usually benefits the borrowers and increased economic activity benefits the bottom-part of income distribution \parencite{furceri2019robust}. Our evidence supports the view of inflation enhancing income inequality, at least up to a threshold, as its square term is borderline on the inclusion, but negative in terms of posterior mean of its coefficient. This would suggest that the above-mentioned theoretical effects interact and manifest with different strength at varying inflation levels.

  The index of the economic freedom of the world describes the overall business conditions by considering the regulatory and legal environment. More economic freedom to trade and run daily business makes it easier to exploit economic opportunities and the high \ac{PIP} with positive posterior mean of the coefficient point to economic freedom making the distribution of income less equal. We consider many other potential explanatory variables, but in the case of the Gini index, they show low \acp{PIP}. We do not find any measure of globalization, institutions, or trade openness relevant to the overall distribution of income. In the next section, we concentrate on the top income shares and whether the top part of the distribution where we observe partially distinct results.

  \subsection{Top income shares}
  Financial development may influence various parts of the income distribution differently. We therefore follow by baseline estimates with the top income shares as dependent variables. Table \ref{res:baseline_top10} reports the results for top 10\% share and Table \ref{res:baseline_top1} for the top 1\% share. There is large overlap of the most important determinants for the top 10\% income share and Gini coefficient. GDP per capita, life expectancy, government expenditures, education index (but not expenditures), and inflation remain among the relevant regressors and their posterior means are consistent with the estimation for the overall income distribution. However, \acp{PIP} of some of the variables decline significantly and some other become relevant. Figure \ref{fig:comp_ineq_sel} provides a visual comparison of the inclusion probabilities.
